---
title: "Intermediate R. Cheatsheet elaborated from datacamp course."
output:
  html_notebook: default
  pdf_document: default
---

# Lest's loop!

### Build a while loop from scratch


```{r}
# Initialize i as 1 
i <- 1

# Code the while loop
while (i <= 10) {
  print(3 * i)
  if (3*i %% 8 == 0) {
    break
  }
  i <- i + 1
}
```
### Loop over a vector

```{r}
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Loop version 1

for (element in linkedin) {
    print(element)
}

# Loop version 2
for (i in 1:length(linkedin)) {
    cat(linkedin[i], "\n")
}
```
### Loop over a list

You need double square brackets - [[ ]] - to select the list elements in loop version 2.

```{r}
# The nyc list is already specified
nyc <- list(pop = 8405837, 
            boroughs = c("Manhattan", "Bronx", "Brooklyn", "Queens", "Staten Island"), 
            capital = FALSE)

# Loop version 1
for (element in nyc) {
    print(element)
}

# Loop version 2
for (i in 1:length(nyc)) {
    print(nyc[[i]])
}
```
### Loop over a matrix

```{r}
tic_tac_toe <- matrix(c("X", "O", "X",
                        "O", "X", "O",
                        "X", "O", "X"), 
                      nrow = 3, ncol = 3, byrow = TRUE)

# define the double for loop
for (i in 1:nrow(tic_tac_toe)) {
  for (j in 1:ncol(tic_tac_toe)) {
    print(paste("On row ", i, "and column ", j , "the board contains ", tic_tac_toe[[i,j]]))
  }
}
```
### Mix it up with control flow

```{r}
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Code the for loop with conditionals
for (li in linkedin) {
  if ( li > 10 ) {
    print("You're popular!")
  } else {
    print("Be more visible!")
  }
  print(li)
}
```
### Next, you break it

```{r}
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Adapt/extend the for loop
for (li in linkedin) {
  if (li > 10) {
    print("You're popular!")
  } else {
    print("Be more visible!")
  }
  # Add if statement with break
  if (li > 16) {
    print("This is ridiculous, I'm outta here!")
    break
  } else if (li < 5) {
    print("This is too embarrassing!")
  }
  print(li)
}
```
### Can you write code that counts the number of r's that come before the first u in rquote?

- Initialize the variable rcount, as 0.
- Finish the for loop:
    - if char equals "r", increase the value of rcount by 1.
    - if char equals "u", leave the for loop entirely with a break.
- Finally, print out the variable rcount to the console to see if your code is correct.

```{r}
# Pre-defined variables
rquote <- "r's internals are irrefutably intriguing"
chars <- strsplit(rquote, split = "")[[1]]

# Initialize rcount
rcount <- 0

# Finish the for loop
for (char in chars) {
  if (char == "r") {
      rcount = rcount + 1
  } else if (char == "u") {
      break
  }
}

# Print out rcount
print(rcount)
```

