---
title: "Intermediate R. Cheatsheet elaborated from datacamp course. Functions."
output: html_notebook
---
### Function documentation

The following two commands, do the same:

`help(sd)`

`?sd`

---

`args()`  is a function to learn about the arguments of a function without having to read through the entire documentation.

The Usage section of the documentation includes two versions of the `mean()` function. The first usage, `mean(x, ...)` is the most general usage of the mean function. The 'Default S3 method', however, is:

`mean(x, trim = 0, na.rm = FALSE, ...)`

The `...` is called the _ellipsis_. It is a way for R to pass arguments along without the function having to name them explicitly. 

---

### Writing my own functions

```
my_function <- function(arg1, arg2) {
  body
}
```
Notice that this recipe uses the assignment operator (`<-`) just as if you were assigning a vector to a variable for example. This is not a coincidence. Creating a function in R basically is the assignment of a function object to a variable! In the recipe above, you're creating a new R variable `my_function`, that becomes available in the workspace as soon as you execute the definition. From then on, you can use the `my_function` as a function.

```{r}
# Create a function pow_two()
pow_two <- function(arg) {
    arg ** 2
}

# Use the function
pow_two(12)

# Create a function sum_abs()
sum_abs <- function(arg1, arg2) {
    abs(arg1) + abs(arg2)
}

# Use the function
sum_abs(2, -3)
```

There are situations in which your function does not require an input. Let's say you want to write a function that gives us the random outcome of throwing a fair die:

```
throw_die <- function() {
  number <- sample(1:6, size = 1)
  number
}

throw_die()
```
```{r}
# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# The interpret() can be used inside interpret_all()
interpret <- function(num_views) {
  if (num_views > 15) {
    print("You're popular!")
    return(num_views)
  } else {
    print("Try to be more visible!")
    return(0)
  }
}

# Define the interpret_all() function
# views: vector with data to interpret
# return_sum: return total number of views on popular days?
interpret_all <- function(views, return_sum=TRUE) {
  count <- 0

  for (v in views) {
    count = count + interpret(v)
  }

  if (return_sum == TRUE) {
    return(count)
  } else {
    return(NULL)
  }
}

# Call the interpret_all() function on both linkedin and facebook
interpret_all(linkedin)
interpret_all(facebook)
```
### R Packages

There are basically two extremely important functions when it comes down to R packages:

- `install.packages()`, which as you can expect, installs a given package.
- `library()` which loads packages, i.e. attaches them to the search list on your R workspace.

`ggvis` is for data visualization. It contains functions such as `layer_points`, `scale_nominal` and `add_axis`.

```{r}
library(ggvis)

ggvis(mtcars, ~wt, ~hp)
```

The `require()` function loads packages into your R session just like the `library()` function. the only difference appears when you're trying to load a package that is not yet installed.

Let's say we want to load the `data.table` package, a package to perform data manimulation, but this package is not yet installed.

```{r}
require("data.table")
```

```{r}
result = require("data.table")
result
```
We can execute the `search()` commands in the console to look at the currently attached packages.

```{r}
library(ggplot2)

qplot(mtcars$wt, mtcars$hp)
```
The `library()` and `require()` functions are not very picky when it comes down to argument types: both `library(rjson)` and `library("rjson")` work perfectly fine for loading a package.

## `lapply` function

Before you go about solving the exercises below, have a look at the documentation of the `lapply()` function. The Usage section shows the following expression:

`lapply(X, FUN, ...)`

To put it generally, `lapply` takes a vector or list `X`, and applies the function `FUN` to each of its members. If `FUN` requires additional arguments, you pass them after you've specified `X` and `FUN` (`...`). The output of `lapply()` is a list, the same length as `X`, where each element is the result of applying `FUN` on the corresponding element of `X`.

```{r}
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")

# Split names from birth year
split_math <- strsplit(pioneers, split = ":")
split_math

# Convert to lowercase strings: split_low
split_low <- lapply(split_math, tolower)

# Take a look at the structure of split_low
str(split_low)
```
### `lapply` and anonymous functions
Writing your own functions and then using them inside `lapply()` is quite an accomplishment! But defining functions to use them only once is kind of overkill, isn't it? That's why you can use so-called anonymous functions in R.

Previously, you learned that functions in R are objects in their own right. This means that they aren't automatically bound to a name. When you create a function, you can use the assignment operator to give the function a name. It's perfectly possible, however, to not give the function a name. This is called an anonymous function:

```
# Named function
triple <- function(x) { 3 * x }

# Anonymous function with same implementation
function(x) { 3 * x }

# Use anonymous function inside lapply()
lapply(list(1,2,3), function(x) { 3 * x })
```
### Use `lapply` with additional arguments

`lapply()` provides a way to handle functions that require more than one argument, such as the `multiply()` function:

```
multiply <- function(x, factor) {
  x * factor
}
```

`lapply(list(1,2,3), multiply, factor = 3)`

### Apply functions that return NULL

`lapply` always returns a list, no matter the input.

```
lapply(split_low, function(x) {
  if (nchar(x[1]) > 5) {
    return(NULL)
  } else {
    return(x[2])
  }
})
```
## `sapply` function

`sapply` is the short for _simplify apply_. It manages to convert the result to a one dimensional array, which is a vector.

```{r}
cities <- c("New York", "Paris", "London", "Tokio", "Rio de Janeiro", "Cape Town")
unlist(lapply(cities, nchar))
```
```{r}
sapply(cities, nchar)
```
`sapply` even found a sensible way of naming this vector. But we can choose not to name the output of `sapply`, by setting its `USE.NAMES = FALSE`.
```{r}
sapply(cities, nchar, USE.NAMES = FALSE)
```
What would happen if the function you want to apply over the input, each time returns a vector containing two values instead of one?
```{r}
first_and_last <- function(name) {
  name <- gsub(" ", "", name)
  letters <- strsplit(name, split = "")[[1]]
  c(first = min(letters), last = max(letters))
}

first_and_last("New York")
```
If we use `sapply` over every city name in cities we obtain a matrix, instead of a vector, with 2 rows and 6 columns.
```{r}
sapply(cities, first_and_last)
```
```{r}
# Create a function that returns min and max of a vector: extremes
temp <- list(
  c(3, 7, 9, 6, -1),
  c(6, 9, 12, 13, 5),
  c(4, 8, 3, -1, -3),
  c(1, 4, 7, 2, -2),
  c(5, 7, 9, 4, 2),
  c(-3, 5, 8, 9, 4),
  c(3, 6, 9, 4, 1)
)

extremes <- function(x) {
  c(min = min(x), max = max(x))
}

# Apply extremes() over temp with sapply()
sapply(temp, extremes)

# Apply extremes() over temp with lapply()
lapply(temp, extremes)
```
### `sapply` with functions that return `NULL`
`sapply()` does not simplify the list of `NULL`'s. That's because the 'vector-version' of a list of `NULL`'s would simply be a `NULL`, which is no longer a vector with the same length as the input.

```{r}
# Definition of print_info()
print_info <- function(x) {
  cat("The average temperature is", mean(x), "\n")
}

# Apply print_info() over temp using sapply()
sapply(temp, print_info)

# Apply print_info() over temp using lapply()
lapply(temp, print_info)
```
## `vapply` function

 he function `vapply()` has the following syntax:

`vapply(X, FUN, FUN.VALUE, ..., USE.NAMES = TRUE)`

Over the elements inside `X`, the function `FUN` is applied. The `FUN.VALUE` argument expects a template for the return argument of this function `FUN`. `USE.NAMES` is `TRUE` by default; in this case `vapply()` tries to generate a named array, if possible.

- explicitly specify output format.

```{r}
sapply(cities, nchar)
```
In our example, we want to apply the `nchar` function over cities. `nchar` is a function that returns a single number, which is a numeric vector of length 1. We can template this output using the `numeric()` function, by setting `FUN.VALUE` to `numeric(1)`, which tells the `vapply` function that `nchar()` should return a single numerical value. The result is exactly the same as the `sapply` function from before. However, this 'pre-specification' of `FUN`'s return value makes `vapply` a safer alternative to `sapply`.

```{r}
vapply(cities, nchar, FUN.VALUE = numeric(1))
```
The following is an example that throws an error and then we fix it not to throw it again.
```{r}
basics <- function(x) {
  c(min = min(x), mean = mean(x), median = median(x), max = max(x))
}

# Fix the error:
vapply(temp, basics, numeric(3))
```
```{r}
# Fix the error:
vapply(temp, basics, numeric(4))
```
---
## Useful functions

`seq` generates a sequence of numbers. The fisrt two arguments tell R the limits of the sequence. That is, where to str and end the sequence respectively.

```{r}
seq1 <- seq(8, -2, by = -2)
seq1
```
The `rep` function has the ability to replicate its input, which tipically is a vector or a list. Using the `times` argument, we can specify how the replication should happen.
```{r}
rep_seq <- rep(seq1, times = 2)
rep_seq
```
We can also use the each argument inside rep
```{r}
rep_each2 <- rep(seq1, each = 2)
rep_each2
```
Then we can `sort` the vector...
```{r}
sort(rep_each2, decreasing = FALSE)
```
The `str()` function inspect the structure of a list.
```{r}
str(seq1)
str(seq)
```
Other useful functions...
```{r}
is.list(seq1)
is.vector(seq1)
```
```{r}
list_seq1 <- as.list(seq1)
is.list(list_seq1)
```
And then we can convert it again to a vector...
```{r}
vec_seq1 <- unlist(list_seq1)
vec_seq1
is.vector(vec_seq1)
```
The `append` function allows you to add elements to a vector or a list in a very readable way.
```{r}
vec_seq1 <- append(vec_seq1, 88)
vec_seq1
```
The `rev` function reverses elements in a data structure.
```{r}
rev_vec_seq1 <- rev(vec_seq1)
rev_vec_seq1
```
### Regular Expresions and other functions

With the `grepl` functions we can determine, which word, for example, has an _a_ in their name.
```{r}
animals <- c("cat", "moose", "impala", "ant", "kiwi")
grepl(pattern = "a", x = animals)
```
For strings that start with an _a_...
```{r}
grepl(pattern = "^a", x = animals)
```
... and at the end fo the word... 
```{r}
grepl(pattern = "a$", x = animals)
```
`grep` function returns a vector of indices of the elements of `x` that yield a match.
```{r}
grep(pattern = "a", x = animals)
```
Now, to replace these matches with other strings we can use `sub`.
```{r}
repla1 <- sub(pattern = "a", replacement = "o", x = animals)
repla1
```
The `sub` function only looks for the first match in the string, and if it finds it, replaces it with the replacement argument and immediately stops looking. If we want to replace every single match of a pattern in a string with the replacement argument, you should try the `gsub` function instead. Now, _impala_ gets converted to _impolo_, so the two "a"'s have been replaced.
```{r}
animals
repla2 <- gsub(pattern = "a", replacement = "o", x = animals)
repla2
```
```{r}
# The emails vector has already been defined for you
emails <- c("john.doe@ivyleague.edu", "education@world.gov", "dalai.lama@peace.org",
            "invalid.edu", "quant@bigdatacollege.edu", "cookie.monster@sesame.tv")

# Use grepl() to match for .edu addresses more robustly
grepl(pattern = "@.*\\.edu$", emails)

# Use grep() to match for .edu addresses more robustly, save result to hits
hits <- grep(pattern = "@.*\\.edu$", emails)

# Subset emails using hits
emails[hits]
```
### Times and Dates

How to deal with times and dates in R...

```{r}
# Ask R what the current date is.
today <- Sys.Date()
today
```
```{r}
# What class is `today`variable?
class(today)
```
Now let's have a look to the time function.
```{r}
now <- Sys.time()
now
```
```{r}
class(now)
```
The important class here is `POSIXct`. Apart from providing rich functionality for calculus and formatting, this class makes sure that the dates and times in R are compatible acrross different operating systems, according to the `POSIX` standard.

Now let's create a date object!
Be aware that `.Date` tries some default data formats, so we have to put de date in and specific order.
```{r}
# For example write the date November 12, 1992
my_wdate <- as.Date("12-11-1992")
my_wdate
```
The correct way to do that would be to put the date as `"1992-11-12"`, but we cannot always do that, especially if we have to many dates to convert.
So, in order to specify the function we have to specify the order in which they are. Examples are listed below.

`%Y`: 4-digit year (1982),\
`%y`: 2-digit year (82),\
`%m`: 2-digit month (01),\
`%d`: 2-digit day of the month (13),\
`%A`: weekday (Wednesday),\
`%a`: abbreviated weekday (Wed),\
`%B`: month (January),\
`%b`: abbreviated month (Jan)\

```{r}
my_date <- as.Date("12-11-1992", format = "%d-%m-%Y")
my_date
```
## Create and format times

Similar to working with dates, you can use `as.POSIXct()` to convert from a character string to a `POSIXct` object, and `format()` to convert from a `POSIXct` object to a character string. Again, you have a wide variety of symbols:

`%H`: hours as a decimal number (00-23),\
`%I`: hours as a decimal number (01-12),\
`%M`: minutes as a decimal number,\
`%S`: seconds as a decimal number,\
`%T`: shorthand notation for the typical format `%H:%M:%S`,\
`%p`: AM/PM indicator\

```{r}
# For checking the full documentation in the console.
?strptime
```
Again, `as.POSIXct()` uses a default format to match character strings. In this case, it's `%Y-%m-%d %H:%M:%S`. In this exercise, abstraction is made of different time zones.

# Excercises

1. Three character strings representing dates have been created for you. Convert them to dates using `as.Date()`, and assign them to `date1`, `date2`, and `date3` respectively. The code for `date1` is already included.
1. Extract useful information from the dates as character strings using `format()`. From the first date, select the weekday. From the second date, select the day of the month. From the third date, you should select the abbreviated month and the 4-digit year, separated by a space.
```{r}
# Definition of character strings representing dates
str1 <- "May 23, '96"
str2 <- "2012-03-15"
str3 <- "30/January/2006"

# Convert the strings to dates: date1, date2, date3
date1 <- as.Date(str1, format = "%b %d, '%y")
date2 <- as.Date(str2, format = "%Y-%m-%d")
date3 <- as.Date(str3, format = "%d/%B/%Y")


# Convert dates to formatted strings
format(date1, "%A")
format(date2, "%d")
format(date3, "%b %Y")
```
1. Convert two strings that represent `timestamps`, `str1` and `str2`, to `POSIXct` objects called `time1` and `time2`.
1. Using `format()`, create a string from `time1` containing only the minutes.
1. From `time2`, extract the hours and minutes as `"hours:minutes AM/PM"`. Refer to the assignment text above to find the correct conversion symbols!
```{r}
# Definition of character strings representing times
str1 <- "May 23, '96 hours:23 minutes:01 seconds:45"
str2 <- "2012-3-12 14:23:08"

# Convert the strings to POSIXct objects: time1, time2
time1 <- as.POSIXct(str1, format = "%B %d, '%y hours:%H minutes:%M seconds:%S")
time2 <- as.POSIXct(str2)

# Convert times to formatted strings
format(time1, "%M")
format(time2, "%I:%M %p")
```

