---
title: "Intermediate R. Cheatsheet elaborated from datacamp course. Functions."
output: html_notebook
---
### Function documentation

The following two commands, do the same:

`help(sd)`

`?sd`

---

`args()`  is a function to learn about the arguments of a function without having to read through the entire documentation.

The Usage section of the documentation includes two versions of the `mean()` function. The first usage, `mean(x, ...)` is the most general usage of the mean function. The 'Default S3 method', however, is:

`mean(x, trim = 0, na.rm = FALSE, ...)`

The `...` is called the _ellipsis_. It is a way for R to pass arguments along without the function having to name them explicitly. 

---

### Writing my own functions

```
my_function <- function(arg1, arg2) {
  body
}
```
Notice that this recipe uses the assignment operator (`<-`) just as if you were assigning a vector to a variable for example. This is not a coincidence. Creating a function in R basically is the assignment of a function object to a variable! In the recipe above, you're creating a new R variable `my_function`, that becomes available in the workspace as soon as you execute the definition. From then on, you can use the `my_function` as a function.

```{r}
# Create a function pow_two()
pow_two <- function(arg) {
    arg ** 2
}

# Use the function
pow_two(12)

# Create a function sum_abs()
sum_abs <- function(arg1, arg2) {
    abs(arg1) + abs(arg2)
}

# Use the function
sum_abs(2, -3)
```

There are situations in which your function does not require an input. Let's say you want to write a function that gives us the random outcome of throwing a fair die:

```
throw_die <- function() {
  number <- sample(1:6, size = 1)
  number
}

throw_die()
```
```{r}
# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# The interpret() can be used inside interpret_all()
interpret <- function(num_views) {
  if (num_views > 15) {
    print("You're popular!")
    return(num_views)
  } else {
    print("Try to be more visible!")
    return(0)
  }
}

# Define the interpret_all() function
# views: vector with data to interpret
# return_sum: return total number of views on popular days?
interpret_all <- function(views, return_sum=TRUE) {
  count <- 0

  for (v in views) {
    count = count + interpret(v)
  }

  if (return_sum == TRUE) {
    return(count)
  } else {
    return(NULL)
  }
}

# Call the interpret_all() function on both linkedin and facebook
interpret_all(linkedin)
interpret_all(facebook)
```
### R Packages

There are basically two extremely important functions when it comes down to R packages:

- `install.packages()`, which as you can expect, installs a given package.
- `library()` which loads packages, i.e. attaches them to the search list on your R workspace.

`ggvis` is for data visualization. It contains functions such as `layer_points`, `scale_nominal` and `add_axis`.

```{r}
library(ggvis)

ggvis(mtcars, ~wt, ~hp)
```

The `require()` function loads packages into your R session just like the `library()` function. the only difference appears when you're trying to load a package that is not yet installed.

Let's say we want to load the `data.table` package, a package to perform data manimulation, but this package is not yet installed.

```{r}
require("data.table")
```

```{r}
result = require("data.table")
result
```
We can execute the `search()` commands in the console to look at the currently attached packages.

```{r}
library(ggplot2)

qplot(mtcars$wt, mtcars$hp)
```
The `library()` and `require()` functions are not very picky when it comes down to argument types: both `library(rjson)` and `library("rjson")` work perfectly fine for loading a package.

### `lapply` function

Before you go about solving the exercises below, have a look at the documentation of the `lapply()` function. The Usage section shows the following expression:

`lapply(X, FUN, ...)`

To put it generally, `lapply` takes a vector or list `X`, and applies the function `FUN` to each of its members. If `FUN` requires additional arguments, you pass them after you've specified `X` and `FUN` (`...`). The output of `lapply()` is a list, the same length as `X`, where each element is the result of applying `FUN` on the corresponding element of `X`.

```{r}
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")

# Split names from birth year
split_math <- strsplit(pioneers, split = ":")
split_math

# Convert to lowercase strings: split_low
split_low <- lapply(split_math, tolower)

# Take a look at the structure of split_low
str(split_low)
```




